Investigating the Effect of Dataset Representivity for Image Colourisation

### 2019-10-12 Paper/paper.tex
Fixed grammar

Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general. To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise grayscale images of a South African context. There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs. This did lead to the hues of the images being slightly more South African in essence, but which was largely unnoticed by the human eye. Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.

@@

Such a phenomenon exists when training a model to recognise and colourise aspects of grayscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.

The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being used initially, particularly a non-South African dataset being used to train a model to operate on a South African sample.

Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset. However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required. Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled or require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.

Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset is such that bias is minimised and the model performs optimally for both sets of images.

@@

For simlicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space. This makes it convenient when seperating the gray and colour channels due to the fact that the 'L' channel in LAB is simply the grayscale image. This grayscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image. These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively. HSB would be another option when choosing a suitable colour space. In general, the choice of colour space makes no difference to the resulting image that our network produces.

@@

When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem. Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland. This is the result of the multi-modality of pictures. There is no clear colour for any one image. This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.

@@

For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand. We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour. If we aimed to produce images which were more diverse in colour and saturated, regression would not suffice. Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the ouput and the representivity of local images in the data, which it does.

@@

Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al. \cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colorization'\cite{zhang2016colorful} approached colourisation as a classification problem.

@@

Our original dataset is a subset of the Places365 dataset. This dataset consists of images that are mainly of European or Western origin. This means that if we train a model on such a dataset and then prodceed to evaluate it on South African or local images, we expect an increase in our loss. This drop in performance can be attributed to the difference in colour of local images. For example, South African landscapes have a distinct hue that is not present in European countries. South African people also have a different appearance to western people who will likely be present in European datasets.

@@

Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a high inclusion rate to the above mentioned epoch. For this reason, we chose to limit our inclusion of South African data to 30\%. Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.

For the sake of simplicity, validation and test sets have been merged into one. For each dataset, there are 40000 images in the training set and 1000 in the test/validation split. This makes it an average sized data set which means that the inclusion rates on these datasets shoud reflect to other datasets of similar size with only minor deviation.

@@

We found that the loss dropped immediately to a reasonable value within the first five epochs but the images produces were not of a high quality. The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile. Better results could have been achieved by training more models, each with a different inclusion of local data. This would allow us to draw a definative correlation of local data to loss but in this case, the samples we have taken will give us a very rough idea of the optimal parameters.

@@

After all models have been trained on their respective datasets, we evaluate them on purely the South African dataset. This gives us a good understanding of how much local data to include into foreign datasets if we want reasonable results. The pure South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split. These were both used to construct the hybrid datasets. Now, we evaluate the model that has been previously trained on the foreign-local hybrid set on a purely South African dataset and measure its performance.

@@

After evalutating each model on the same purely South African dataset, these are their respective losses,

@@

As you can see, the diffrence in loss is miniscule. The small drop in loss that you get when including local data is not worth the time and effort spent retrieving, processing and preparing it in the first place. The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant. By no means does this mean that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin. It does, however, mean that the inclusion of local data simply does not effect the outcome of the experiment as much as other factors would. Other factors include running the experiment for longer or simply finding a larger dataset. The insignificance of local data inclusion is made extremely evident when comparing images that were produced using a model trained without any local data with a model trained using 30\% local data,

@@

Due to the nature of our experiment, the amount of local inclusion did not dramatically effect the outcome, but that would not be the case when dealing with facial recognition as previously discussed. In that case, it would be vital to have an inclusive dataset that is well distributed across all races and genders.

@@

Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye and so does not justify the additional resources required to build the inclusive dataset.

@@

Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect. Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is does solely by computers with a report merely being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.

>>>>

Sufficient research exists for the implementation of bias-mitigating machine learning algorithms, however, the presence of bias in the original dataset has not been comprehensively considered, which could lead to vast improvements in the efficiency and diversity of machine learning in general. To determine what percentage of local data has to be included in order to alleviate this bias, we examine the sub-problem of how much South African data is required to correctly colourise greyscale images of a South African context. There was only a 3.3\% reduction in the error of the colourised images using separate regression-based models trained to 75 epochs with 0\% to 30\% inclusion of local data, with minimal reduction in loss occurring after 5 epochs. This did lead to the hues of the images being slightly more South African in essence, but which was largely unnoticed by the human eye. Consequently, the inclusion of local data in the dataset definitely minimises the bias of the dataset, but whether or not that bias-mitigation justifies the vast resources required to build the comprehensive dataset is determined on a case-by-case basis.

@@

Such a phenomenon exists when training a model to recognise and colourise aspects of greyscale images: by employing a training dataset with a bias to a particular locale – like an American or a Eurocentric setting, for instance – any sample taken from outside such an area (with particular focus placed on a South African sample) is misrepresented and yields erroneous results, with the produced colourised image being quintessentially different to the intended image.

The identified cause is the lack of diversity in the training datasets because even though various methods can be implemented in the actual model to mitigate this bias, they cannot account for a wholly prejudiced dataset being initially used, particularly a non-South African dataset being used to train a model to operate on a South African sample.

Therefore, the aim is to investigate what percentage of the training dataset has to be of local origin compared to the international origin of the remainder of the dataset. However, constructing and testing datasets to determine such a goal is both time consuming and expensive in terms of financial and computational resources required. Hence, the approach of image colourisation has been selected as it allows an uncomplicated procedure of collecting publicly-available images of both International and South African settings, thus providing a simple way to build a diverse dataset which does not need to be labelled nor does it require excessive computational time since the machine learning approach required is one of supervised learning where the expected outcome is derived from the images themselves.

Consequently, the refined aim is to investigate the optimal ratio of non-South African images to South African images in the training dataset such that bias is minimised and the model performs optimally for both sets of images.

@@

For simplicity, rather than using the typical RGB colour space, we make use of the CIELAB colour space. This makes it convenient when separating the grey and colour channels due to the fact that the 'L' channel in LAB is simply the greyscale image. This greyscale input (1 x W x H) is then passed through the model and compared with the output (2 x W x H) of the original 'AB' channels of the image. These 'AB' channels make up the colour and simply signify the red-green and blue-yellow shift respectively. HSB would be another option when choosing a suitable colour space. In general, the choice of colour space makes no difference to the resulting image that our network produces.

@@

When colourising images using a deep learning approach, it is generally easiest to develop the model as a regression problem. Classification would yield better results which are more life-like and saturated whereas regression would lead to output images appearing dull and bland. This is the result of the multi-modality of pictures: there is no clear colour for any one image. This leads our model, using an aggregate loss function, to simply find an average colour, resulting in a bland image.

@@

For the sake of simplicity, we continue to use regression as it leads to satisfactory results whilst being simple to understand. We utilize a mean squared error loss which may average colour distribution but still produces images with a reasonable amount of colour. If we aimed to produce images which are more diverse in colour and saturated, regression would not suffice. Our model should, however, be able to compare output across datasets in order to draw correlation between the accuracy of the output and the representivity of local images in the data, which it does.

@@

Our choice of model has been based on the simple regression model developed by Luke Melas \cite{melas2018mark} and loosely based on the network used by Zhang et al. \cite{zhang2016colorful}\cite{zhang2017real} with the major difference being that their model used in 'Colorful Image Colourisation'\cite{zhang2016colorful} approached colourisation as a classification problem.

@@

Our original dataset is a subset of the Places365 dataset. This dataset consists of images that are mainly of European or Western origin. This means that if we train a model on such a dataset and then proceed to evaluate it on South African or local images, we expect an increase in our loss. This drop in performance can be attributed to the difference in colour of local images. For example, South African landscapes have a distinct hue that is not present in European countries. South African people also have a different appearance to Western people, who are likely be present in European datasets.

@@

Due to the fact that images collected from Flickr were substantially larger in size than those of the European set, it took longer to train models with a higher inclusion rate to the above mentioned epoch. For this reason, we chose to limit our inclusion of South African data to a maximum of 30\%. Future experiments could show that an inclusion rate higher than 30\% is optimal but for the purpose of this experiment, it will suffice.

For the sake of simplicity, validation and test sets have been merged into one. For each dataset, there are 40000 images in the training set and 1000 in the test/validation split. This makes it an average sized data set which means that the inclusion rates on these datasets should reflect to other datasets of similar size with only minor deviation.

@@

We found that the loss dropped immediately to a reasonable value within the first five epochs but that the images produced were not of a high quality. The training process was long and slow with very little change in loss from one epoch to the next but the end results make it worthwhile. Better results could have been achieved by training more models, each with a different inclusion of local data. This would allow us to draw a definitive correlation of local data to loss but in this case, the samples we have taken will give us an approximate idea of the optimal parameters.

@@

After all models have been trained on their respective datasets, we evaluate them on the purely South African dataset. This gives us a good understanding of how much local data to include with foreign datasets if we want reasonable results. The solely South African dataset contains 35000 images in its training split and 5000 images in the validation/testing split. These were both used to construct the hybrid datasets. Now, we evaluate the model that has been previously trained on the foreign-local hybrid set on a purely South African dataset and measure its performance.

@@

After evaluating each model on the same purely South African dataset, these are their respective losses:

@@

Clearly, the difference in loss is minuscule. The small drop in loss that you get when including local data does not warrant the time and effort spent retrieving, processing and preparing at first. The majority of datasets used in image colourisation or classification already contain a wide variety of images that will make including local data redundant. By no means does this imply that the inclusion of local data plays no role in producing better looking images that more closely resemble the place of origin. It does, however, mean that the inclusion of local data merely does not effect the outcome of the experiment as much as other factors would. Other factors include running the experiment for longer or simply finding a larger dataset. The insignificance of local data inclusion is made extremely evident when comparing images that were produced using a model trained without any local data with a model trained using 30\% local data,

@@

@@

Consequently, the effect of the inclusion of local data is not as profound as one might have hypothesised and the images produced from a model trained with this local data included are largely indistinguishable from the images produced from a corresponding international dataset, especially to the human eye, and so does not justify the additional resources required to build the inclusive dataset.

@@

Although we investigated the effect of dataset bias on image colourisation, there are a multitude of other applications of machine learning where this bias may come into effect. Due to the inherent differences between the nature of image colourisation and other applications (for example, in examining and predicting economic trends in a particular city), where humans may be abstracted from the analysis of the data and it is done solely by computers, with a corresponding report being generated - in such a case, minute changes are of the utmost importance and lead to significant changes in the market trends recognised.
